---
title: "Data wrangling, part II"
output:
  html_document:
    css: lab.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
library(knitr)
library(tidyverse)
library(skimr)
opts_chunk$set(eval=TRUE)
```


Today's wrangling exercises are modified from [Master the Tidyverse](https://github.com/rstudio-education/master-the-tidyverse).

## Single-table verbs

The `dplyr` package is where the most common data wrangling "verbs" come from. The main ones are:

- `select()` to pick out variables/columns
- `filter()` to pick out cases/rows
- `arrange()`, which arranges cases
- `summarize()`, a many-to-few or many-to-one operation to suummarize
- `mutate()`, a many-to-many operation to add additional variables. 
- `group_by()`, a helper to create grouped data frames

We'll focus primarily on `filter()`, `summarize()`, `group_by()`, and `mutate()`. For this section,  we'll be working with the `babynames` data, which comes in its own package

```{r}
# install.packages("babynames")
library(babynames)
head(babynames)
tail(babynames)
```

This dataset includes baby name data from the US Social Security Administration, for all names which had at least 5 uses in a particular year. 

### Filter

Recall that we can use `filter()` to exctract just the rows we're interested in, as in

```{r}
filter(babynames, name == "Amelia")
```

Other than `==`, you can use other logical tests, such as `<`, `>`, `<=`, `>=`, `!=` (not equal) and `%in%` (group membership).

See if you can use the logical operators to find

1. all of the names where `prop` is greater than or equal to 0.08
2. all of the children named "Sea"

You can combine several criteria together using Boolean operators, such as `&` (and) and `|` (or). Commas also work in `filter()` for a list of criteria that must all be met. 

```{r}
filter(babynames, name=="Amelia", year==1880) #this is equivalent to
filter(babynames, name=="Amelia" & year == 1880)
```

Try to find 

3. All girls named "Sea"
4. Names that were used by exactly 5 or 6 children in 1880
5. Names that are one of Acura, Lexus, or Yugo
6. The row that corresponds to you (or, a good friend, if you have an uncommon name)

### Summarize

Summarise is a many-to-one or many-to-few operation. It reduces the size of your data by summarizing it. 

```{r}
babynames %>% 
  dplyr::summarize(total = sum(n), max = max(n))
```

Use summarize to compute:

7. The first (minimum) year in the dataset
8. The last (maximum) year in the dataset

Now, put together your `filter()` and `summarize()` skills to extract the rows where name == "Khaleesi" and find

9. The total number of children named Khaleesi
10. The first year Khaleesi appeared in the data

### Group by

We've seen `group_by()` in an incidental way many times this semester. `group_by()` is a `dplyr` verb that creates a grouped data frame. It can be combined with other `dplyr` verbs very powerfully. 

```{r}
babynames %>%
  group_by(sex)

babynames %>%
  group_by(sex) %>%
  dplyr::summarize(total = sum(n))
```

### Mutate

Mutate is the way to make a new variable in a dataset. It is a many-to-many operation. 

```{r}
babynames %>%
  mutate(percent = round(prop*100, 2))
```

Of course, you would need to use the assignment operator or another pipe to do something with this new variable. 


## Multiple table verbs

If you have more than one dataset, you may want to do a "join" to connect them together in some way. This is where multiple-table verbs come in. In `dplyr`, they are `left_join()`, `right_join()`, `full_join()`, `inner_join()`, `semi_join()` and `anti_join()`. We will focus primarily on the `left_join()` because I find it to be useful the most often in data wrangling.

For this section, we'll use a baby example about bands to illustrate how these joins differ.

```{r}
band <- tribble(
  ~name,  ~band,
  "Mick", "Stones",
  "John", "Beatles",
  "Paul", "Beatles"
)
instrument <- tribble(
  ~name, ~plays,
  "John", "guitar",
  "Paul", "bass",
  "Keith", "guitar"
)

instrument2 <- tribble(
  ~artist, ~plays,
  "John", "guitar",
  "Paul", "bass",
  "Keith", "guitar"
)
```

### Left join

```{r}
band %>%
  left_join(instrument, by = "name")
```


### Right join

```{r}
band %>%
  right_join(instrument, by = "name")
```


9. How do the results of the left and right joins differ?

### Full join

```{r}
band %>%
  full_join(instrument, by = "name")
```

### Inner join

```{r}
band %>%
  inner_join(instrument, by = "name")
```

### What if the names don't match?

```{r}
band %>% 
  left_join(instrument2, by = c("name" = "artist"))
```

### Filtering joins

The joins we've looked at so far are mutating joins, but there are also filtering joins, which use information from one data set to extract cases from another data set. 

Semi-join

```{r}
band %>% 
  semi_join(instrument, by = "name")
```

Anti-join
```{r}
band %>% 
  anti_join(instrument, by = "name")
```

### Practice

For this section, we'll use data from the `nycflights13` package, which you will need to install.

```{r}
# install.packages("nycflights13")
library(nycflights13)
```

The package has multiple datasets, with different sets of variables. The main dataset is named `flights`.

```{r}
flights
```

But, the variable `carrier` doesn't make much sense to people outside the airline industry. So, we'd like to join on some easier-to-understand names, which are in the `airlines` dataset.

```{r}
airlines
```

10. Which airlines had the largest arrival delays? Work in groups to complete the code below.

```{r, eval=FALSE}
flights %>%
  drop_na(arr_delay) %>%
                     %>% # add some joining code here
  group_by(        ) %>% # add a variable here
                     %>% # compute the average arrival delay
  arrange(         ) # arrange by delays
```

11. How many airports in airports are serviced by flights originating in New York (i.e. flights in our dataset?) Notice that the column to join on is named `faa` in the `airports` dataset and `dest` in the `flights` dataset.
