---
title: "ggplot2 and aesthetics"
output: html_document
---

```{r steup, message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
```

If you don't define aesthetics within the overall `ggplot` call, `geom_`s won't know what to do. 

For example, this will give an error, because the `geom_smooth` has no variables to work with:
```{r, error=TRUE}
ggplot(data=mtcars) + geom_point(aes(x=cyl, y=mpg)) + geom_smooth(method="lm", se=FALSE)
```


This works, but is verbose, because it repeats the same thing several times. 
```{r}
ggplot(data=mtcars) + geom_point(aes(x=cyl, y=mpg)) + geom_smooth(aes(x=cyl, y=mpg), method="lm", se=FALSE)
```


There's a shortcut, though. When you write ggplot2 code, the *aes*thetics in the `ggplot` call get applied everywhere down the chain. 
This allows you to do more compact things like this: 
```{r}
ggplot(data=mtcars, aes(x=cyl, y=mpg)) + geom_point() + geom_smooth(method="lm", se=FALSE)
```

But, if you want to add coloring, it can make things weird if you stick them into the overall `ggplot` aesthetics.

```{r}
mtcars <- mtcars %>%
  mutate(vs=factor(vs)) #factors are R's way of dealing with categorical variables

ggplot(data=mtcars, aes(x=cyl, y=mpg, col=vs)) + geom_point() + geom_smooth(method="lm", se=FALSE)
```

Whoops, now we have two lines! If we want the color to only happen to one geom, we can specify it just in that one.

```{r}
ggplot(data=mtcars, aes(x=cyl, y=mpg)) + geom_point(aes(col=vs)) + geom_smooth(method="lm", se=FALSE)
```