## Your turn 1

Look at the R object WorldPhones (by typing its name in your notebook or the Console and hitting enter). 

```{r}

```


## Vectors and matrices

```{r}
vec <- c(1, 2, 3, 10, 100)
vec
```

```{r}
mat <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 2)
mat

matrix(c(1, 2, 3, 4, 5, 6), nrow = 2)
matrix(c(1, 2, 3, 4, 5, 6), nrow = 3)
```

## Math with vectors and matrices

```{r}
vec + 4
vec * 4
vec * vec # element-wise multiplication

vec %*% vec # matrix multiplication (inner)
vec %o% vec # matrix multiplication (outer)

mat
t(mat) # transpose
```

## Arrays

```{r}
array(c(1, 2, 3, 4, 5, 6), dim = c(2, 2, 3))
```

## Data types
Numeric
```{r}
1 + 1
3000000
class(0.00001)
```

Character
```{r}
"hello"
class("hello")
```

```{r, error=TRUE}
"hello" + "world"
```

```{r}
nchar("hello")
paste("hello", "world")
```

Logical
```{r}
3 > 4
class(TRUE)
class(T)
```

Factor (danger zone!)
```{r}
fac <- factor(c("a", "b", "c"))
fac
class(fac)
```

## Your turn 

Make a vector that contains the number 1, the letter R, and the logical TRUE

```{r}

```

## Quiz
What type of data will result?

## Lists

```{r}
lst <- list(1, "R", TRUE)
class(lst)
```

## Data frames
```{r}
df <- data.frame(c(1, 2, 3), 
 c("R","S","T"), c(TRUE, FALSE, TRUE))
class(df)
```

## Naming 
```{r}
nvec <- c(one = 1, two = 2, three = 3)
nlst <- list(one = 1, two = 2, many = c(3, 4, 5))
```

```{r}
ndf <- data.frame(numbers = c(1, 2, 3), 
          letters = c("R","S","T"), 
          logic = c(TRUE, FALSE, TRUE))
```

```{r}
names(ndf)
names(nvec)
```



## Subsetting

## The dplyr way
To use the dplyr package, we need to load it.

```{r}
library(dplyr)
```



## Beatles data
Let's make a toy dataset to play with. 

```{r}
beatles <- data.frame(
  name = c("John", "Paul", "George", "Ringo"),
  birth = c(1940, 1942, 1943, 1940), 
  instrument = c("guitar", "bass", "guitar", "drums")
)
```

## Select
Select is a way to extract columns from dataframes

```{r}
select(beatles, name, birth)
```

## Your turn
Select just the instrument column



## Base R
```{r}
beatles[2,3]
beatles[ ,"birth"]
beatles[ ,c("name","birth")]
beatles$birth
```

## Your turn
Select just the instrument column using brackets

```{r}

```

Select just the instrument column using a dollar sign.

```{r}

```

## Filter

```{r}
filter(beatles, name == "George")
```

## Your turn


## More than one condition
```{r}
filter(beatles, birth==1940, instrument == "guitar")
filter(beatles, birth==1940 & instrument == "guitar")
```

## Your turn
Modify  the code below to filter out the rows for which birth is 1943 or instrument is drums

```{r}
filter(beatles, birth==1940 & instrument == "guitar")
```


## Base R

```{r}
beatles[c(FALSE,TRUE,TRUE,FALSE), ]
beatles[beatles$birth == 1940, ]
```

## Your turn

