Except for the group roster, all submissions will be made via Canvas. 

          Checkpoint           Due Date    Credit   File name
  -------------------------- ------------- -------- ----------------------------------
         Group roster            10/15        1 pt    STAT 320 Roster
   Initial project proposal      10/29       10 pts  STAT 320 Group-X Initial Proposal
    Revised project design       11/19       10 pts  STAT 320 Group-X Revised Proposal
        Data appendix            11/26       15 pts   STAT 320 Group-X Data Appendix      
      Oral presentation         12/13        30 pts  STAT 320 Group-X Presentation
        Project report           12/21      40 pts  STAT 320 Group-X Project Report
        Group Dynamic            12/21       5 pts   STAT 320 Group-X Dynamic

<!--
        Draft handout            4/28      10 pts  SDS 291 Group-X Draft Handout



Date   | Groups
-------|--------
Dec 8  | K, I, G, J, L, D
Dec 10 | C, B, F, M, A, E, H
Dec 16 | Technical Report, Group Dynamic due

--> 

Group order for presentations
```{r, eval=FALSE, echo=FALSE}
set.seed(417)
groups <- sample(LETTERS[1:10], replace = FALSE)
groups
``` 

```{r, eval=FALSE, echo=FALSE}
groups = c("L", "M", "N", "O", "P", "Q", "R", "S", "T")
shuffle(groups)
``` 