---
title: "Introduction to R and RStudio"
author: "Your Name"
date: "2/3/2017"
---

```{r, message=FALSE}
require(mosaic)
require(oilabs)
```

In the previous few pages, you recreated some of the displays and preliminary analysis of Arbuthnot's baptism data. Your assignment involves repeating these steps, but for present day birth records in the United States. Load up the present day data with the following command.

```{r}
data(present)
```

The data are stored in a data frame called `present`.

##### Question 1

What years are included in this data set? What are the dimensions of the data frame and what are the variable or column names?

```{r}

```

##### Question 2

How do the counts of boys and girls in the present day birth records compare to Arbuthnot's? Are they on a similar scale?

```{r}

```

##### Question 3

Make a plot that displays the boy-to-girl ratio for every year in the data set. What do you see? Does Arbuthnot’s observation about boys being born in greater proportion than girls hold up in the U.S.? Include the plot in your response.

```{r}

```

##### Question 4

In what year did we see the most total number of births in the U.S.?

```{r}

```
