\documentclass[10pt]{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{fancyhdr,url}
\usepackage{graphicx}

\oddsidemargin 0in  %0.5in
\topmargin     0in
\leftmargin    0in
\rightmargin   0in
\textheight    8.9in
\textwidth     6in %6in
%\headheight    0in
%\headsep       0in
%\footskip      0.5in

\newtheorem{thm}{Theorem}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{obs}{Observation}
\newtheorem{lemma}{Lemma}
\newtheorem{claim}{Claim}
\newtheorem{definition}{Definition}
\newtheorem{question}{Question}
\newtheorem{answer}{Answer}
\newtheorem{problem}{Problem}
\newtheorem{solution}{Solution}
\newtheorem{conjecture}{Conjecture}

\pagestyle{fancy}

\lhead{\textsc{Prof. McNamara}}
\chead{\textsc{SDS/MTH 220: Presentations}}
\rhead{\textsc{12/11/2017-12/13/2017}}
\lfoot{}
\cfoot{\thepage}
\rfoot{}
\renewcommand{\headrulewidth}{0.2pt}
\renewcommand{\footrulewidth}{0.0pt}

\begin{document}
%\maketitle

\newcommand{\grademe}[2]{
\begin{table}[h!]
	\centering
	\begin{tabular}{|l|c|c|c|c|c|p{2.5in}|}
	\hline
	Criteria & Poor & Fair & Average & Good & Excellent & Comments \\
	\hline
	Presentation format 		& 1 & 2 & 3 & 4 & 5 & \\[1.5ex]
	Engagement	& 1 & 2 & 3 & 4 & 5 & \\[1.5ex]
	Professionalism	& 1 & 2 & 3 & 4 & 5 & \\[1.5ex]
	Presence 	& 1 & 2 & 3 & 4 & 5 & \\[1.5ex]
	Content 		& 1 & 2 & 3 & 4 & 5 & \\[1.5ex]
	\hline
	\end{tabular}
	\caption{Group #1 - #2}
\end{table}
}


\paragraph{Peer Evaluation}

Your assistance is required in evaluating the presentations of your peers. Please the circle the number that best describes your perception of each group's presentation. Keep the following grading rubrics in mind. 

\begin{itemize}
	\item Presentation format: Were the slides readable? Could you follow the narrative of their presentation?
	\item Engagement: Do the speakers seem to be interested in the material they are presenting? Are they explaining their topic in such a way it engages you?
	\item Professionalism: Do the speakers use intonation to signal what words or concepts are most important? Did they appear to have practiced ahead of time? Was it clear they had divided the presentation between them? Did the speakers stay on time? 
	\item Presence: Do the speakers command the room? Do they draw your attention or do they just look like they want this to be over as soon as possible? Are the speakers loud enough? Can you hear them even in the back of the room?  Are the speakers engaged with the audience? Do they seem connected or do they seem to be going through the motions?	
	\item Content: Do the speakers appear to have full command of the material? Were there obvious gaps in their knowledge of the subject matter? Were they able to answer questions coherently? 
\end{itemize}

\grademe{A}{Students}
%\grademe{C}{Charlotte, Heather, Cadence, Sarah}
%\grademe{K}{Ghida, Jenny}
%\grademe{B}{Pam, Sophia, Celia}
%\grademe{A}{Jenny, Lisa, Mikayla}
%\grademe{H}{Natalie, Veida, Daisy}
%\grademe{E}{Leen, Maya, Cecilia}
%\grademe{D}{Linnea, Anna, Sarah}
%\grademe{J}{Nicole, Ilana, Dana}
%\grademe{G}{Deepal, Anna, Laura}
%\grademe{I}{Nicole, Aissah, Ruby}
%\grademe{M}{Ambar, Jess}
%\grademe{F}{Margaret, Serena, Mandira}
%\grademe{L}{Jakqueline, Rutendo, Lizette}



\end{document}
