\documentclass[10pt]{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{fancyhdr}
\usepackage{graphicx}

\oddsidemargin 0in  %0.5in
\topmargin     0in
\leftmargin    0in
\rightmargin   0in
\textheight    9in
\textwidth     6in %6in
%\headheight    0in
%\headsep       0in
%\footskip      0.5in

\newtheorem{thm}{Theorem}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{obs}{Observation}
\newtheorem{lemma}{Lemma}
\newtheorem{claim}{Claim}
\newtheorem{definition}{Definition}
\newtheorem{question}{Question}
\newtheorem{answer}{Answer}
\newtheorem{problem}{Problem}
\newtheorem{solution}{Solution}
\newtheorem{conjecture}{Conjecture}

\pagestyle{fancy}

\lhead{\textsc{Prof. McNamara}}
\chead{\textsc{STAT 320: Pre-Course Questionnaire}}
\rhead{\textsc{Fall 2018}}
\lfoot{}
\cfoot{}
%\cfoot{\thepage}
\rfoot{}
\renewcommand{\headrulewidth}{0.2pt}
\renewcommand{\footrulewidth}{0.0pt}



\begin{document}
%\maketitle

\paragraph*{Purpose} 
Thank you for completing this questionnaire, which is helpful to me in structuring the course.  All information will be kept confidential.

\begin{center}
\begin{tabular}{|c|c|c|c|c|}
	\hline
	Full name & Prefer to be called & Pronouns & Email address \\
	\hline
	\hspace{1.75in} & \hspace{1in} & \hspace{0.5in} & \hspace{2.15in} \\[2ex]
	\hline
\end{tabular}
\newline
\vspace*{0.1in}
\newline
\begin{tabular}{|c|c|c|}
	\hline
	Major & Class year & Registration status (circle one) \\
	\hline
	\hspace{3in} & \hspace{0.9in} & enrolled / wait list / neither \\ [2ex]
	\hline
\end{tabular}
\end{center}

\subsubsection*{Background}
\begin{itemize}
  \itemsep0.9in
	\item Please tell me a little about your background (e.g. major(s) and area of interest, other activities, significant orgs or athletics):
	\item Please list the three most recent quantitative courses you've taken, (which may have been in high school):
	\item This course uses R for statistical computing. Have you ever used R before? Do you know any other programming languages?
	\item What are your personal goals for this course?
%	\item Are there specific topics that you are particularly interested in studying?
	\item In which areas of applied statistics are you most interested? (e.g. biology, economics, psychology, finance, sociology, etc.) List as many as you like. 
	\vspace{0.7in}
%	\item How strongly do you entertain the notion of working in a quantitative field after graduation (circle one)?
%	\begin{center}
%		not at all \qquad hardly ever \qquad sometimes \qquad often \qquad OMG, it's all I think about!
%	\end{center}
  \vspace{-0.7in}
	\item Is there any other information that I should know?
\end{itemize}




\end{document}
